SPENCER THESIS - WORK IN PROGRESS

### 2016-06-03 doc/thesis_swe.tex
Thesis update. #98

Guided by these results, in this thesis I build upon the work of Collet et al. in \cite{collet} to apply a simple dark matter halo model to reconstruct lensing mass along lines of sight in the Millennium Simulation in order to predict the weakly lensed ellipticities of generated background sources. I then compare my predicted lensed ellipticities to the `true' lensing calculated by Hilbert et al. in \cite{ray_tracing} by ray-tracing light through the Millennium Simulation. By doing so, I attempt to answer the following questions: Under the best possible circumstances of having noiseless redshifts and halo masses, how accurately can I reconstruct line-of-sight dark matter mass structures using a halo model? What fraction of dark matter mass is in halos? How feasible is an inference of the Millennium Simulation's cosmological hyper-parameters, such as the stellar mass - halo mass relation, using all available halo data? How long does such an inference take?

%Guided by these results, in this thesis I build upon the work of Collet et al. in \cite{collet} to apply a simple dark matter halo model to reconstruct lensing mass along lines of sight in the Millennium Simulation using the \texttt{Pangloss} framework in order to predict the weakly lensed ellipticities of generated background sources.

A brief introduction to the halo model, the effects of strong and weak gravitational lensing, representing galaxy ellipticities, and cosmological correlation functions is discussed in Section \ref{background}. The implementation of the \texttt{Pangloss} framework to make line of sight lensing predictions in the Millennium Simulation is described in Section \ref{pangloss}, and the results of the framework on ?? galaxies on a field size of ?? arcmin$^2$, along with a comparison of the predicted lensed ellipticities to ray-traced ellipticities, is given in Section \ref{check}. Section \ref{feasibility} explores the feasibility of using the presented \texttt{Pangloss} framework to make parameter inferences using all available halo data either by likelihood estimation or approximate Bayesian computation. Section \ref{discussion} discusses limitations of the used framework as well as potential physical and computational improvements that can be made for upcoming work before concluding remarks in Section \ref{conclusion}.

@@

\begin{align}\label{gg_corr_def}
\xi_\pm(\Delta\theta)&=\left<\varepsilon_{i_t}(\vec{\theta})\varepsilon_{j_t}(\vec{\theta}+\vec{\Delta\theta})\right>\pm\left<\varepsilon_{i_\times}(\vec{\theta})\varepsilon_{j_\times}(\vec{\theta}+\vec{\Delta\theta})\right>\\
\xi_\times(\vec{\Delta\theta})&=\left<\varepsilon_{i_t}(\vec{\theta})\varepsilon_{j_\times}(\vec{\theta}+\vec{\Delta\theta})\right>
\end{align}

\noindent where $\vec{\theta}$ is the position vector of the first galaxy and $\Delta\vec{\theta}$ is the angular separation vector between the pair of galaxies \cite{schneider}. Gravitational lensing shears ellipticities tangentially, so $\xi_\times(\Delta\theta)$ should vanish with enough galaxies and therefore is a useful estimate of the bias. (\textcolor{red}{Why is $\xi_-$ expected to be zero?}), which leaves $\xi_+(\Delta\theta)$ as the desired probe of lensing mass.

As galaxies will only be lensed by the same foreground mass at relatively small separation distances, the $\varepsilon$-$\varepsilon$ correlation function decreases sharply as the separation increases. If the distribution of intrinsic galaxy ellipticities is indeed random, then their $\varepsilon$-$\varepsilon$ correlation function should vanish across all separation scales.

@@

%  and is defined as
% \begin{equation}\label{ng_corr_def}
% \xi=
% \end{equation}

% \noindent where...\textcolor{red}{Can't seem to find definition??}

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Generating Background Galaxies}
With the (0,0,0,0) catalog of MS foreground galaxies chosen, a set of ?? background galaxies over a field of ?? arcmin$^2$ was generated, with density 10 galaxies per arcmin$^2$. The intrinsic orientation of each galaxy was sampled from a uniform distribution as, without lensing, there should be no preferred orientation due to the assumption of an isotropic universe. The magnitude of the galaxy ellipticities was sampled from a normal distribution with a standard deviation of 0.2, but any ellipticities with magnitude greater than one were re-sampled. Random ellipticity noise was sampled from a normal distribution with a standard deviation of 0.1 and added to the intrinsic ellipticities. Finally, each ellipticity was multiplied by $M=0.9$ to account for a multiplicative bias of 10\%.

@@

In this work, the threshold values used are ${M_T=10^{12}M_\odot}$ and ${R_T=10}$ kpc. The relevance distribution for one background catalog realization with ?? foreground halos in a lightcone of radius ?? arcminutes is shown in Figure \ref{fig:rel_dist} As expected, there are very few massive halos along the line of sight and therefore only a small percentage of foreground halos should have a high relative contribution to the predicted lensing quantities. This encouraged the intended analysis on lightcones with all halos and those with only a small subset of the most relevant halos to compare the results and see if the computational time saved by using only the most relevant halos was worth the loss in prediction accuracy.

@@

@@

Using the presented methodology, ??? background galaxies in a ??? arcmin$^2$ subset of the (0,0,0,0) Millennium Simulation foreground catalog were generated and lensed by both the ray-traced shear and convergence maps as well as the \texttt{Pangloss} framework. With a lightcone radius of ?? arcminutes, the lightcones contained an average of ${??\pm??}$ foreground halos. Both sets of lensed ellipticities were characterized with correlation functions, as well as the intrinsic ellipticities of the sources before lensing.

@@

The first test of the halo model was with the $\varepsilon$-$\varepsilon$ correlation function with a lightcone radius of ?? arcminutes which is given in Figure \ref{gg_corr}. The statistic measured the average correlation between pairs of ellipticities at separation distances between ?? and ?? arcminutes. The cross-component of the correlation function shows no significant deviation from zero as expected, as gravitational lensing only shears galaxies tangentially. For the $\xi_+$ component, the ray-traced (plotted in green) and halo model (plotted in purple) are in relative agreement from seperation scales of ?? to ?? minutes, but there is a clear systematic underprediction of shear power by the halo model on scales larger than ??. This may be indicative that the halo model does not adequately address large-scale mass structures, or that there is significant dark matter mass not correlated with galaxies. The normalized root-mean-square error (NRMSE) of the halo correlation is ??.

Using the same catalog of background sources, this statistic was recalculated for the halo model ellipticities for various lightcone radii ranging from $R_L=1$ to $R_L=8$ arcminutes. The result is shown in Figure \ref{gg_corr_series}, where the series of model predicted correlation functions is compared to the ray-traced correlation function. As $R_L$ increases, the mean number of foreground galaxies contained in each lightcone grows quadratically from ${??\pm??}$ galaxies when $R_L=1$ arcminute to ${??\pm??}$ galaxies when $R_L=8$ arcminutes. Increasing the number of foreground objects, and thus increasing the mass and structure considered for shear and convergence predictions, systematically increased the predicted shear power on all scales.

%While the series of predicted correlation functions appear to be converging towards the ray-traced correlation function, there is clearly diminishing shear power gains as $R_L$ increases.

@@

The catalog of lensed ellipticities was also analyzed using the galaxy-mass correlation function with separation distances from ?? to ?? arcminutes and a lightcone radius of ?? arcminutes. The result is shown in Figure \ref{ng_corr}. The lensed ellipticities from the ray-tracing and halo model both show positive correlation on all calculated scales and are in agreement on separations from ?? to ?? arcminutes. However, the halo model is still systematically underpredicting the shear power on scales above ?? arcminutes, although now only by an average of ??\%. A new feature is the large overprediction of shear power on separation scales smaller than ?? arcminutes. This is likely because background sources this close to foreground halos are often in the strong lensing regime which is not currently accounted for by the \texttt{Pangloss} framework. The NRMSE of the halo correlation is ??.

The galaxy-mass correlation function was similarly re-calculated at increasing lightcone radii from $R=1$ to $R=8$ arcminutes with the result in Figure \ref{ng_corr_series} As with the $\varepsilon$-$\varepsilon$ correlation, ...

@@

Of intrinsic interest is the relative number of relevant halos needed in each lightcone to accurately predict the halo correlation functions in Figures ?? and ??, as a low percentage would allow much faster lensing predictions. Figure ?? displays an example ellipticity-ellipticity correlation function comparrison of the halo model using all halos (purple) and halo model using halos above the relevance limit of (?? $M_{\odot}/$Mpc$^3$), as well as the percent error of each predicted value between the two implementations. This relevance limit resulted in $??\pm??$ relevant halos per lightcone, and resulted in a normalized root-mean-square error (NRMSE) of ??. Figures ?? plots the ellipticity-ellipticity and galaxy-mass correlation function NRMSE and mean correlation fraction as a function of mean relevant halos. These results suggest that either correlation function prediction can be approximated with an NRMSE below ?? using the ?? most relevant halos. (More analysis after plots have been made)

%%------------------------------------------------------------------------------------------------------------------------------------------
% Correlation Function Comparrison
\begin{figure}
    \centering
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/rel_gg_compare.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        %The ellipticity-ellipticity correlation function for ?? sources at separation distances from 0.1 to 2 arcminutes and a lightcone radius of $R=2$ arcminutes. While the correlation of the intrinsic ellipticities before lensing (blue) shows no correlation, the correlation of the ray-traced ellipticities (green) and halo model predicted ellipticities (purple) both show significant positive correlation on scales up to 1 arcminute. The halo model predicts a small negative correlation above this scale, while the ray-tracing shows positive correlation on all scales.}
        %The correlation of the intrinsic ellipticities before lensing is plotted in blue, and shows no correlation. The correlation of the true ray-traced ellipticities is plotted in green and shows significant positive correlation on all scales. The halo predicted correlation, plotted in purple, shows significant positive correlation until a separation distance of 0.8 arcminutes.
        \label{fig:rel_gg_compare}
    \end{subfigure}
    ~~
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/rel_ng_compare.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        %The ellipticity-ellipticity correlation functions for a series of halo model frameworks with increasing lightcone radii from $R=1$ to $R=8$ arcminutes, all plotted in purple with various line styles. These are compared to the same ray-traced correlation from Figure \ref{gg_corr}.}
        \label{fig:rel_ng_compare}
    \end{subfigure}
    \caption{}
\end{figure}

@@

%------------------------------------------------------------------------------------------------------------------------------------------
% NRMSE and Correlation Fraction as function of Mean Relevant Halos
% \begin{figure}
%     \centering
%     \includegraphics[width=0.475\textwidth]{figs-swe/thesis/corr_fraction.png}
%     \captionsetup{justification=raggedright,singlelinecheck=false}
%     \caption{CAPTION}
%     \label{corr_fraction}
% \end{figure}

@@

(Analyze profile for future optimization recommendations!!)

@@

(Introduce Bayes' Theorem, likelihood, prior, posterior, etc.??)

@@

A possible implemenetation of the above for \texttt{Pangloss} would be as follows: Choose the the ray-traced lensed ellipticities as the `observed' data and compute its $\varepsilon$-$\varepsilon$ correlation function to be used as the desired summary statistic. Let the distance metric to be the NRMSE between the predicted and ray-traced $\varepsilon$-$\varepsilon$ correlation function. Compute the best-case $\varepsilon$-$\varepsilon$ correlation function using all halo data as shown in Section \ref{pangloss} and make the threshold value $\delta$ be some multiple of the best-case NRMSE (e.g. ${\delta=2\cdot\text{RMSE}_{\text{best-case}}}$). Create mock data by repeating the methodology in \ref{pangloss} except use the stellar mass - halo mass relation, dependent on paramters we sample from a prior, to predict halo masses based upon each galaxy's stellar mass. Compute the simulated $\varepsilon$-$\varepsilon$ correlation function and determine whether to accept or reject the data by determining if the NRMSE is below the threshold $\delta$. The posterior will be estimated after a large number of iterations through the rejection algorithm.

@@

I implemented the methodology in section \ref{Pangloss} to make weakly lensed ellipticity predictions in 10 mock catalogs of $N$ background sources ranging from $N=??$ to $N=??$. The mean CPU times for a single likelihood calulation and ABC rejection decision as a function of $N$ is shown in Figure \ref{mle_vs_abc}. (Clearly abc is slower, and inherently approximate, and requires arbitrary threshold cutoff, so clearly we will be using MLE moving forward.) Additionally, the time spent for either inference method is quite small; the CPU time per likelihood estimation or ABC rejection decision are \textit{far} less than the CPU time required to create lightcones and make the LOS lensing prediction. This is an encouraging result and suggests that further work on the framework should be on additional optimization of the lensing prediction code rather than posterior estimation.

%%------------------------------------------------------------------------------------------------------------------------------------------
% ABC Diagram
% \begin{figure}
%     \centering
%     \includegraphics[width=0.475\textwidth]{figs-swe/thesis/mle_vs_abc.png}
%     \captionsetup{justification=raggedright,singlelinecheck=false}
%     \caption{CAPTION}
%     \label{fig:mle_vs_abc}
% \end{figure}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Issues with the Halo Model}

\begin{enumerate}
\item Under-predicting shear on large scales
\item Not accounting for strong lensing
\item Smooth-component correction not enough
\end{enumerate}

% OLD: Both the ellipticity-ellipticity and galaxy-mass correlation functions show clear issues with using the halo model to predict the lensing of background sources with \texttt{Pangloss}, even in the best-case scenario of perfect knowledge of the halo masses and their spectroscopic redshifts. The systematic underprediction of shear power present in both correlation functions is likely do to missing features in the model such as clusters, filaments, and voids.

@@

% The progression of model predicted ellipticity-ellipticity correlation functions at larger radii converging to the ray-traced correlation in Figure \ref{gg_corr_series} supports the analysis given above, as the prediction was progressively better with larger number of foreground halos and thus more mass and structure. However, the number of foreground objects scales quadratically with the lightcone radius making it prohibitively expensive above $R=8$ arcminutes. With a radius this large, most foreground objects will be too far away from the line of sight and have insufficient mass to make a significant contribution to the lensing of the background source. This suggests the creation of a metric to predict the `relevance' of each halo to the lensing contribution. In this way, the framework could be restructured to only use the most important halos in the computationally expensive lensing prediction, while simultaneously allowing for larger lightcones with more relevant halos. This could aid in the aforementioned shear power prediction issues.

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Scaling Up Pangloss}

While the current \texttt{Pangloss} framework can handle the lensing predictions of ?? sources in a ?? arcmin$^2$ field comfortably on a single processor, the goal is to scale up the framework to make predictions for background sources of the same number density in a 100 deg$^2$ field; this will require 3.6 million lightcones, each containing at least ??? relevant foreground halos. Luckily, the prediction is trivially parallelizable as the shear and convergence calculation of each lightcone is independent of all other lightcones. This makes GPUs an attractive candidate for future work, as it would only take 360 GPUs with 10,000 threads each to carry out the prediction. Additionally, the ?? MB of RAM per lightcone is sufficiently small to fit 10,000 lightcones on a GPU.\\

(More??)

@@

The results of Section \ref{feasibility} demonstrated that the likelihood calculation does not present a computational issue in comparrison to the lensing calculation, so MLE is a feasible approach for a hierarchial inference of the stellar mass - halo mass relation. A single ABC rejection algorithm in fact performed \textit{worse} than a single likelihood calculation, so clearly ABC is not suitable for this work. However as discussed above, there is still much more work to the \texttt{Pangloss} framework that needs to be done, both in modeling features and computational infastructure, before an inference using a large number of background sources comparable to modern sky surves can be achieved.

@@

(Might need to rework some phrasing)\\

>>>>

Guided by these results, in this thesis I build upon the work of Collet et al. in \cite{collet} to apply a simple dark matter halo model to reconstruct lensing mass along lines of sight in the Millennium Simulation in order to predict the weakly lensed ellipticities of generated background sources. I then compare my predicted lensed ellipticities to the `true' lensing calculated by Hilbert et al. in \cite{ray_tracing} by ray-tracing light through the Millennium Simulation. By doing so, I attempt to answer the following questions:

\begin{enumerate}
\item Under the best possible circumstances of having noiseless redshifts and halo masses, how accurately can I reconstruct line-of-sight dark matter mass structures using a halo model?
%\item What fraction of dark matter mass is in halos?
\item How many halos are needed to make these predictions?
\item How feasible is an inference of the Millennium Simulation's cosmological hyper-parameters, such as the stellar mass - halo mass relation, using all available halo data? 
\item How long does such an inference take?
\end{enumerate}
%Guided by these results, in this thesis I build upon the work of Collet et al. in \cite{collet} to apply a simple dark matter halo model to reconstruct lensing mass along lines of sight in the Millennium Simulation using the \texttt{Pangloss} framework in order to predict the weakly lensed ellipticities of generated background sources.

A brief introduction to the halo model, the effects of strong and weak gravitational lensing, representing galaxy ellipticities, and cosmological correlation functions is discussed in Section \ref{background}. The implementation of the \texttt{Pangloss} framework to make line of sight lensing predictions in the Millennium Simulation is described in Section \ref{pangloss}, and the results of the framework on 1440 galaxies on a field size of 144 arcmin$^2$, along with a comparison of the predicted lensed ellipticities to ray-traced ellipticities, is given in Section \ref{check}. Section \ref{feasibility} explores the feasibility of using the presented \texttt{Pangloss} framework to make parameter inferences using all available halo data either by likelihood estimation or approximate Bayesian computation. Section \ref{discussion} discusses limitations of the used framework as well as potential physical and computational improvements that can be made for upcoming work before concluding remarks in Section \ref{conclusion}.

@@

\begin{align}\label{gg_corr_def}
\xi_\pm(\Delta\theta)&=\left<\varepsilon_{i_t}(\vec{\theta})\,\varepsilon_{j_t}(\vec{\theta}+\vec{\Delta\theta})\right>\nonumber \\
&\pm\left<\varepsilon_{i_\times}(\vec{\theta})\,\varepsilon_{j_\times}(\vec{\theta}+\vec{\Delta\theta})\right>\\
\xi_\times(\vec{\Delta\theta})&=\left<\varepsilon_{i_t}(\vec{\theta})\,\varepsilon_{j_\times}(\vec{\theta}+\vec{\Delta\theta})\right>
\end{align}

\noindent where $\vec{\theta}$ is the position vector of the first galaxy and $\Delta\vec{\theta}$ is the angular separation vector between the pair of galaxies \cite{schneider}. Gravitational lensing shears ellipticities tangentially, so $\xi_\times(\Delta\theta)$ should vanish with enough galaxies and therefore is a useful estimate of the bias. The amplitude of $\xi_-$ is significantly lower than $\xi_+$ as it depends on the initial orientation of \textit{both} ellipticities and not just their offset, which leaves $\xi_+(\Delta\theta)$ as the desired probe of lensing mass. As galaxies will only be lensed by the same foreground mass at relatively small separation distances, the $\varepsilon$-$\varepsilon$ correlation function decreases sharply as the separation increases. If the distribution of intrinsic galaxy ellipticities is indeed random, then their $\varepsilon$-$\varepsilon$ correlation function should vanish across all separation scales.

@@

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Generating Background Galaxies}
With the (0,0,0,0) catalog of MS foreground galaxies chosen, a set of 1440 background galaxies over a field of 144 arcmin$^2$ was generated, with density 10 galaxies per arcmin$^2$. The intrinsic orientation of each galaxy was sampled from a uniform distribution as, without lensing, there should be no preferred orientation due to the assumption of an isotropic universe. The magnitude of the galaxy ellipticities was sampled from a normal distribution with a standard deviation of 0.2, but any ellipticities with magnitude greater than one were re-sampled. Random ellipticity noise was sampled from a normal distribution with a standard deviation of 0.1 and added to the intrinsic ellipticities. Finally, each ellipticity was multiplied by $M=0.9$ to account for a multiplicative bias of 10\%.

@@

In this work, the threshold values used are ${M_T=10^{12}M_\odot}$ and ${R_T=10}$ kpc. The relevance distribution for one background catalog realization with 1440 foreground halos in a lightcone of radius 6 arcminutes is shown in Figure \ref{fig:rel_dist} As expected, there are very few massive halos along the line of sight and therefore only a small percentage of foreground halos should have a high relative contribution to the predicted lensing quantities. This encouraged the intended analysis on lightcones with all halos and those with only a small subset of the most relevant halos to compare the results and see if the computational time saved by using only the most relevant halos was worth the loss in prediction accuracy.

@@

\textcolor{red}{Talk about lensing table!!! ??}

@@

Using the presented methodology, 1440 background galaxies in a 144 arcmin$^2$ subset of the (0,0,0,0) Millennium Simulation foreground catalog were generated and lensed by both the ray-traced shear and convergence maps as well as the \texttt{Pangloss} framework. With a lightcone radius of 6 arcminutes, the lightcones contained an average of ${4836\pm104}$ foreground halos. Both sets of lensed ellipticities were characterized with correlation functions, as well as the intrinsic ellipticities of the sources before lensing.

@@

The first test of the halo model was with the $\varepsilon$-$\varepsilon$ correlation function with a lightcone radius of 6 arcminutes which is given in Figure \ref{gg_corr}. The statistic measured the average correlation between pairs of ellipticities at separation distances between 0.1 and 2 arcminutes. The cross-component of the correlation function shows no significant deviation from zero as expected, as gravitational lensing only shears galaxies tangentially. For the $\xi_+$ component, the ray-traced (plotted in green) and halo model (plotted in purple) are in relative agreement from seperation scales of 0.1 to 0.2 arcminutes, but there is a clear systematic underprediction of correlation by the halo model on scales larger than 0.2 arcminutes. This may be indicative that not enough halo mass is being used in the lensing prediction, the halo model does not adequately address large-scale mass structures, or that there is significant dark matter mass not correlated with galaxies. These issues are discussed further in Section \ref{issues}.

Using the same catalog of background sources, this statistic was recalculated for the halo model ellipticities for various lightcone radii ranging from $R_L=1$ to $R_L=8$ arcminutes. The result is shown in Figure \ref{gg_corr_series}, where the series of model predicted correlation functions is compared to the ray-traced correlation function. As $R_L$ increases, the mean number of foreground galaxies contained in each lightcone grows quadratically from ${134\pm17}$ galaxies when $R_L=1$ arcminute to ${8570\pm87}$ galaxies when $R_L=8$ arcminutes. Increasing the number of foreground objects, and thus increasing the mass and structure considered for shear and convergence predictions, systematically increased the predicted correlation on all scales. This result encourages the use of lightcones with radii of at least 8 arcminutes, and further emphasizes the need for computational efficiency to allow the timely use of large lightcones.

@@

The catalog of lensed ellipticities was also analyzed using the galaxy-mass correlation function with separation distances from 0.1 to 2 arcminutes and a lightcone radius of 6 arcminutes. The result is shown in Figure \ref{ng_corr}. The lensed ellipticities from the ray-tracing and halo model both show positive correlation on all calculated scales and largely agree on scales larger than 0.3 arcminutes except for a slight overprediction by the halo model near 0.45 arcminutes. However, a new feature is the large overprediction of correlation on separation scales smaller than 0.2 arcminutes. This is likely because background sources this close to massive foreground halos are often in the strong lensing regime which is not currently accounted for by the \texttt{Pangloss} framework.

The galaxy-mass correlation function was similarly re-calculated at increasing lightcone radii from $R=1$ to $R=8$ arcminutes with the result in Figure \ref{ng_corr_series} As with the $\varepsilon$-$\varepsilon$ correlation, structure larger than $\sim$1 arcminute is captured much better by increasingly large lightcone radii as expected. However, there is little to no improvement of the galaxy-mass correlation on scales smaller than 0.3 arcminutes which futher points to the need to model strong lensing.

@@

Of intrinsic interest is the relative number of relevant halos needed in each lightcone to accurately predict the halo correlation functions in Figures \ref{fig:gg_corr} and \ref{fig:ng_corr}, as a low percentage would allow much faster lensing predictions. Figure \ref{fig:rel_corr} displays an example $\varepsilon$-$\varepsilon$ and galaxy-mass correlation function comparrison of the halo model using all halos (purple) and halo model using halos above the relevance limit of ${10^{-5} M_{\odot}/\text{Mpc}^3}$, as well as the percent error of each predicted value between the two implementations. This relevance limit resulted in $??\pm??$ relevant halos per lightcone, and resulted in a normalized root-mean-square error (NRMSE) of ??. Figure \ref{fig:halo_fraction} plots the $\varepsilon$-$\varepsilon$ and galaxy-mass correlation function NRMSE as a function of relevance limits and mean relevant halos. These results suggest that either correlation function prediction can be approximated with an NRMSE below 10\% using around the ?? most relevant halos, or an NRMSE of below 5\% using around the ?? most relevant halos. Calibration curves such as this can allow the best-case correlation function to be predicted within a desired accuracy at only a fraction of the full lensing calculation which will be especially helpful in future analyses with much larger fields of view.

%%------------------------------------------------------------------------------------------------------------------------------------------
% Correlation Function Comparrison
\begin{figure}
    \centering
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/rel_gg_compare.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        %The ellipticity-ellipticity correlation function for ?? sources at separation distances from 0.1 to 2 arcminutes and a lightcone radius of $R=2$ arcminutes. While the correlation of the intrinsic ellipticities before lensing (blue) shows no correlation, the correlation of the ray-traced ellipticities (green) and halo model predicted ellipticities (purple) both show significant positive correlation on scales up to 1 arcminute. The halo model predicts a small negative correlation above this scale, while the ray-tracing shows positive correlation on all scales.}
        %The correlation of the intrinsic ellipticities before lensing is plotted in blue, and shows no correlation. The correlation of the true ray-traced ellipticities is plotted in green and shows significant positive correlation on all scales. The halo predicted correlation, plotted in purple, shows significant positive correlation until a separation distance of 0.8 arcminutes.
        \label{fig:rel_gg_compare}
    \end{subfigure}
    ~~
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/rel_ng_compare.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        %The ellipticity-ellipticity correlation functions for a series of halo model frameworks with increasing lightcone radii from $R=1$ to $R=8$ arcminutes, all plotted in purple with various line styles. These are compared to the same ray-traced correlation from Figure \ref{gg_corr}.}
        \label{fig:rel_ng_compare}
    \end{subfigure}
    \caption{}
    \label{fig:rel_corr}
\end{figure}

@@

%------------------------------------------------------------------------------------------------------------------------------------------
% NRMSE and Correlation Fraction as function of Mean Relevant Halos
\begin{figure}
    \centering
    \includegraphics[width=0.475\textwidth]{figs-swe/thesis/halo_fraction.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{CAPTION!}
    \label{fig:halo_fraction}
\end{figure}

@@

The distribution of total CPU time in significant parts of the lensing calculation for run 6 are as follows: 24.0\% of total time calculating individual halo convergence contributions where 14.4\% is spent in lensing lookup tables, 65.1\% of CPU time spent combining all convergence contributions where 62.1\% is spent on the smooth-component correction, and the majority of the remaining time spent on writing \texttt{Astropy}\footnote{http://www.astropy.org/} and \texttt{NumPy}\footnote{http://www.numpy.org/} columns.

While progress has been made, the code needs to be further optimized to allow lensing predictions of fields of many square degrees to be done in a few seconds. Section \ref{scaling_up} discusses the details of future computational scale-up plans.

@@

@@

A possible implemenetation of the above for \texttt{Pangloss} would be as follows: Choose the the ray-traced lensed ellipticities as the `observed' data and compute its $\varepsilon$-$\varepsilon$ correlation function to be used as the desired summary statistic. Let the distance metric to be the NRMSE between the predicted and ray-traced $\varepsilon$-$\varepsilon$ correlation function. Compute the best-case $\varepsilon$-$\varepsilon$ correlation function using all halo data as shown in Section \ref{pangloss} and make the threshold value $\delta$ be some multiple of the best-case NRMSE (e.g. ${\delta=2\cdot\text{NRMSE}_{\text{best-case}}}$). Create mock data by repeating the methodology in \ref{pangloss} except use the stellar mass - halo mass relation, dependent on paramters we sample from a prior, to predict halo masses based upon each galaxy's stellar mass. Compute the simulated $\varepsilon$-$\varepsilon$ correlation function and determine whether to accept or reject the data by determining if the NRMSE is below the threshold $\delta$. The posterior will be estimated after a large number of iterations through the rejection algorithm.

@@

I implemented the methodology in section \ref{Pangloss} to make weakly lensed ellipticity predictions in 20 mock catalogs of $N$ background sources ranging from $N=10^2$ to $N=10^6$. The mean CPU times for a single likelihood calulation and ABC rejection decision as a function of $N$ is shown in Figure \ref{mle_vs_abc}. The two methods take approximately the same amount of time up to $N=10^3$ sources, but the correlation function quickly becomes far too computationally inefficient at $N$ values above $10^4$. As ABC is fundamentally an approximation of the posterior \textit{and} a single decision is orders of magnitude slower than a likelihood calculation, any future inference work should use MLE instead of ABC. Additionally, the time spent for either inference method is quite small in comparrison to the CPU time required to create lightcones and make the line-of-sight lensing predictions. This is an encouraging result and suggests that further work on the framework should be on additional optimization of the lensing prediction code rather than posterior estimation.

%%------------------------------------------------------------------------------------------------------------------------------------------
% ABC Diagram
\begin{figure}
    \centering
    \includegraphics[width=0.475\textwidth]{figs-swe/thesis/mle_vs_abc.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{CAPTION}
    \label{fig:mle_vs_abc}
\end{figure}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Issues with the Halo Model} \label{issues}

Both the $\varepsilon$-$\varepsilon$ and galaxy-mass correlation functions show clear issues with using the halo model to predict the lensing of background sources with \texttt{Pangloss}, even in the best-case scenario of perfect knowledge of the halo masses and their spectroscopic redshifts. The systematic underprediction of correlation on scales larger than ?? in the $\varepsilon$-$\varepsilon$ function indicates that larger structures like filaments and voids are not being fully captured by the halo model. This is not entierly surprising, as the only structure beyond halos accounted for in the presented model was the smooth-component correction which only adds sheets of uniform mass; clearly a better model of filaments and voids is needed to make accurate correlation predictions on scales larger than $\sim$1 arcminute. The progression of model predicted ellipticity-ellipticity correlation functions at larger radii converging to the ray-traced correlation in Figure \ref{gg_corr_series} supports the analysis given above, as the prediction was progressively better with larger number of foreground halos and thus more mass and structure.

A possible solution, briefly discussed in Section \ref{scc}, would be to model filaments as `cylinders' of dark matter attached to the most massive halos with densities, length distributions, and frequencies matching simulated data. This implementation might lead to better correlation predictions on large scales, but the details of the implementation, as well as analyzing how filaments affect line-of-sight lensing measurements, still needs to be explored. Stellar masses could also be incorporated into the \texttt{Pangloss} prediction, but any effect would likely be dominated by the missing features described above. Resolving the overprediction of correlation on scales below ?? in the galaxy-mass function is likely more straightforward as sources this close to the center of foreground mass are often in the strong lensing regime. \texttt{Pangloss} currently removes any sources with total reduced shear $g$ greater than 0.5, but this potentially leaves sources that are not entierly described by weak lensing. Correctly modeling line-of-sight strong lensing will require analyzing the convergence and shear contributions of BMO halos in the strong lening regime and at the interface between strong and weak lensing.

(Mention something about kappamaps)

@@

%The progression of model predicted ellipticity-ellipticity correlation functions at larger radii converging to the ray-traced correlation in Figure \ref{gg_corr_series} supports the analysis given above, as the prediction was progressively better with larger number of foreground halos and thus more mass and structure. However, the number of foreground objects scales quadratically with the lightcone radius making it prohibitively expensive above $R_L=8$ arcminutes without using only a small fraction of the most relevant halos. %With a radius this large, most foreground objects will be too far away from the line of sight and have insufficient mass to make a significant contribution to the lensing of the background source. This suggests the creation of a metric to predict the `relevance' of each halo to the lensing contribution. In this way, the framework could be restructured to only use the most important halos in the computationally expensive lensing prediction, while simultaneously allowing for larger lightcones with more relevant halos. This could aid in the aforementioned correlation prediction issues.

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Scaling Up Pangloss} \label{scaling_up}

While the current \texttt{Pangloss} framework can handle the lensing predictions of 1440 sources in a 144 arcmin$^2$ field comfortably on a single processor, the goal is to scale up the framework to make predictions for background sources of the same number density in a 100 deg$^2$ field; this will require 3.6 million lightcones, each containing at least ??? relevant foreground halos. Luckily, the prediction is trivially parallelizable as the shear and convergence calculation of each lightcone is independent of all other lightcones. This makes GPUs an attractive candidate for future work, as it would only take 360 GPUs with 10,000 threads each to carry out the prediction. Additionally, the ?? MB of RAM per lightcone is sufficiently small to fit 10,000 lightcones on a GPU.\\

(Future code optimizations to get source prediction under ?? seconds)

@@

The results of Section \ref{feasibility} demonstrated that the likelihood calculation does not present a computational issue in comparrison to the lensing calculation, so MLE is a feasible approach for a hierarchial inference of the stellar mass - halo mass relation. A single ABC rejection algorithm in fact performed \textit{worse} than a single likelihood calculation for ${10^2\leq N\leq10^6}$ sources, so ABC does not appear to be suitable for this work. However as discussed above, there is still much more work to the \texttt{Pangloss} framework that needs to be done, both in modeling features and computational infastructure, before an inference using a large number of background sources comparable to modern sky surves can be achieved.

@@
